---
title: "Cyclistic Bike Share Analysis"
author: "Kade Golden"
date: "2023-02-28"
subtitle: "How Does A Bike Share Navigate Speedy Success?"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
 
  **This project is in relation to the Google Professional Data Anayltics Certificate**
 
  
  You are a data analyst working in the marketing analyst team at Cyclistic, a bike-share company in Chicago. The director of marketing believes the company’s future success depends on maximizing the number of annual memberships. Therefore, your team wants to understand how casual riders and annual members use Cyclistic bikes differently. 
  
  From these insights, your team will design a new marketing strategy to convert casual riders into annual members. But first, Cyclistic executives must approve your recommendations, so they must be backed up with compelling data insights and professional data visualizations.
  
  
**Characters and Teams**

 * Cyclistic: A bike-share program that features more than 5,800 bicycles and 600 docking stations. Cyclistic sets itself apart by also offering reclining bikes, hand tricycles, and cargo bikes, making bike-share more inclusive to people with disabilities and riders who can’t use a standard two-wheeled bike. The majority of riders opt for traditional bikes; about 8% of riders use the assistive options. Cyclistic users are more likely to ride for leisure, but about 30% use them to commute to work each day.

 * Lily Moreno: The director of marketing and your manager. Moreno is responsible for the development of campaigns and initiatives to promote the bike-share program. These may include email, social media, and other channels.

 * Cyclistic marketing analytics team: A team of data analysts who are responsible for collecting, analyzing, and reporting data that helps guide Cyclistic marketing strategy. You joined this team six months ago and have been busy learning about Cyclistic’s mission and business goals — as well as how you, as a junior data analyst, can help Cyclistic achieve them.

 * Cyclistic executive team: The notoriously detail-oriented executive team will decide whether to approve the recommended marketing program.

**Ask** 

How do annual members and casual riders use Cyclistic Bike Share differently?

**Prepare**

You will use Cyclistic’s 12 month historical trip data to analyze and identify trends

**Process**

R studio to process, analyzie, and clean data

### Install packages; tidyverse, lubridate, ggplot, 

```{r}
library(tidyverse)
library(lubridate)
library(ggplot2)
```
### Set the working directory 
```{r}
getwd()
## display your working directory
setwd("~/Downloads/Bikeshare_capstone/")
```

### Upload divvy-tripdata.csv files for past 12 months
```{r}
r_202004 <- df <- read.csv('/Users/kadegolden/Downloads/Bikeshare_capstone/202004-divvy-tripdata.csv', header=TRUE, stringsAsFactors=FALSE)
r_202005 <- df <- read.csv('/Users/kadegolden/Downloads/Bikeshare_capstone/202005-divvy-tripdata.csv', header=TRUE, stringsAsFactors=FALSE)
r_202006 <- df <- read.csv('/Users/kadegolden/Downloads/Bikeshare_capstone/202006-divvy-tripdata.csv', header=TRUE, stringsAsFactors=FALSE)
r_202007 <- df <- read.csv('/Users/kadegolden/Downloads/Bikeshare_capstone/202007-divvy-tripdata.csv', header=TRUE, stringsAsFactors=FALSE)
r_202008 <- df <- read.csv('/Users/kadegolden/Downloads/Bikeshare_capstone/202008-divvy-tripdata.csv', header=TRUE, stringsAsFactors=FALSE)
r_202009 <- df <- read.csv('/Users/kadegolden/Downloads/Bikeshare_capstone/202009-divvy-tripdata.csv', header=TRUE, stringsAsFactors=FALSE)
r_202010 <- df <- read.csv('/Users/kadegolden/Downloads/Bikeshare_capstone/202010-divvy-tripdata.csv', header=TRUE, stringsAsFactors=FALSE)
r_202011 <- df <- read.csv('/Users/kadegolden/Downloads/Bikeshare_capstone/202011-divvy-tripdata.csv', header=TRUE, stringsAsFactors=FALSE)
r_202012 <- df <- read.csv('/Users/kadegolden/Downloads/Bikeshare_capstone/202012-divvy-tripdata.csv', header=TRUE, stringsAsFactors=FALSE)
r_202101 <- df <- read.csv('/Users/kadegolden/Downloads/Bikeshare_capstone/202101-divvy-tripdata.csv', header=TRUE, stringsAsFactors=FALSE)
r_202102 <- df <- read.csv('/Users/kadegolden/Downloads/Bikeshare_capstone/202102-divvy-tripdata.csv', header=TRUE, stringsAsFactors=FALSE)
r_202103 <- df <- read.csv('/Users/kadegolden/Downloads/Bikeshare_capstone/202103-divvy-tripdata.csv', header=TRUE, stringsAsFactors=FALSE)
```

### Check the structure the data
```{r}
str(r_202004)
str(r_202005)
str(r_202006)
str(r_202007)
str(r_202008)
str(r_202009)
str(r_202010)
str(r_202011)
str(r_202012)
str(r_202101)
str(r_202102)
str(r_202103)
```
### Change start_station_id data type to integer
```{r}
r_202012$start_station_id <- as.integer(r_202012$start_station_id)
r_202101$start_station_id <- as.integer(r_202101$start_station_id)
r_202102$start_station_id <- as.integer(r_202102$start_station_id)
r_202103$start_station_id <- as.integer(r_202103$start_station_id)
```

### Change end_station_id data type to integer
```{r}
r_202012$end_station_id <- as.integer(r_202012$end_station_id)
r_202101$end_station_id <- as.integer(r_202101$end_station_id)
r_202102$end_station_id <- as.integer(r_202102$end_station_id)
r_202103$end_station_id <- as.integer(r_202103$end_station_id)
```
### Stack individual month data into one data frame
```{r}
all_trips <- bind_rows(r_202004, r_202005, r_202006, r_202007, r_202008, r_202009, r_202010,
                       r_202011, r_202012, r_202101, r_202102, r_202103)
```
### Preview with glimpse to see the consolidation
```{r}
str(all_trips)
glimpse(all_trips)
```
#### Install dplyr
```{r}
library(dplyr)
```
### Check tables and values
```{r}
table(all_trips$member_casual)
table(all_trips$rideable_type)
```
### Install lubridate package for date cleaning
```{r}
library(lubridate)
```
### Add columns that list the date, month, day, and year of each ride
```{r}
all_trips$date <- as.Date(all_trips$started_at)
all_trips$month <- format(as.Date(all_trips$date), "%m")
all_trips$day <- format(as.Date(all_trips$date), "%d")
all_trips$year <- format(as.Date(all_trips$date), "%Y")
all_trips$day_of_week <- format(as.Date(all_trips$date), "%A")
```
### Add a ride_length calculation to all_trips in minutes
```{r}
all_trips$ride_length <- difftime(all_trips$ended_at, all_trips$started_at)

all_trips$ride_length <- all_trips$ride_length/60

all_trips$ride_length <- round(all_trips$ride_length, 2)
```
### Inspect structure 
```{r}
str(all_trips)
```
### Convert ride_length from factor to numeric for calculations
```{r}
is.factor(all_trips$ride_length)
all_trips$ride_length <- as.numeric(as.character(all_trips$ride_length))
is.numeric(all_trips$ride_length)
```
### Create a new data frame and remove bad data
```{r}
all_trips_v2 <- all_trips[!(all_trips$start_station_name == "HQ QR" | all_trips$ride_length<0),]
```
### Remove NA and " " 
```{r}
all_trips_v2 <- na.omit(all_trips_v2)
```

### Check for test stations " TEST"
```{r}
unique(all_trips_v2$start_station_name[grep("TEST", all_trips_v2$start_station_name)])
```

```{r}
unique(all_trips_v2$start_station_name[grep("test", all_trips_v2$start_station_name)])
```
### Clean dataset for test stations and stations with blank values
```{r}
all_trips_v2 <- (filter(all_trips_v2, !('start_station_name' == "WATSON TESTING -DIVVY"
  | 'start_station_name' =="HUBBARD ST BIKE CHECKING (LBS-WH-TEST)" | 'start_station_name' == "hubbard_test_lws"
  |'start_station_name' == "")))
```
### Descriptive Analysis 
```{r}
mean(all_trips_v2$ride_length) # straight average (total ride lengths/ rides)
```
```{r}
median(all_trips_v2$ride_length) # midpoint number in the ascending array of ride lengths
```
```{r}
max(all_trips_v2$ride_length) #longest ride
```
```{r}
min(all_trips_v2$ride_length) #shortest ride
```

```{r}
summary(all_trips_v2$ride_length)
```
### Compare members and casual users
```{r}
aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual, FUN = mean)
aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual, FUN = median)
aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual, FUN = max)
aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual, FUN = min)
```
### See average ride time by each day for member_casual
```{r}
aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual + all_trips_v2$day_of_week, FUN = mean)
```
### Order days of the week 
```{r}
all_trips_v2$day_of_week <- ordered(all_trips_v2$day_of_week, levels=c("Sunday", "Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday"))
```
### See average ride time by each day for member_casual
```{r}
aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual + all_trips_v2$day_of_week, FUN = mean)
```
### Library
```{r}
library(dplyr)
library(ggplot2)
library(tidyverse)
```
### Take out of scientific notation 
```{r}
options(scipen=999)
```

### Analyze ridership data by type and weekday 
```{r}
all_trips_v2 %>% 
  mutate(weekday = wday(started_at, label = TRUE)) %>% 
  group_by(member_casual, weekday) %>% 
  summarise(number_of_rides = n(),average_duration = mean(ride_length)) %>% 
  arrange(member_casual, weekday)  %>% 
  ggplot(aes(x = weekday, y = number_of_rides, fill = member_casual)) +
  geom_col(position = "dodge") + labs(title="Total Number of Rides by Day", x = "Week Day", y = "Number of Rides") + theme(axis.text.x = element_text(angle = 60, hjust = 1))
```


### Average duration of ride for member_casual
```{r}
all_trips_v2 %>% 
  mutate(weekday = wday(started_at, label = TRUE)) %>% 
  group_by(member_casual, weekday) %>% 
  summarise(number_of_rides = n()
            ,average_duration = mean(ride_length)) %>% 
  arrange(member_casual, weekday)  %>% 
  ggplot(aes(x = weekday, y = average_duration, fill = member_casual)) +
  geom_col(position = "dodge") + labs(title="Average Duration of Ride", x = "Week Day", y = "Average Duration Minutes") + theme(axis.text.x = element_text(angle = 60, hjust = 1))

```


### View ridership by type and month 
```{r}
all_trips_v2 %>% 
  group_by(member_casual, month) %>% 
  summarise(number_of_rides = n(), average_duration = mean(ride_length)) %>% 
  arrange(member_casual, month)	%>% 
  ggplot(aes(x = month, y = number_of_rides, fill = member_casual)) +
  geom_col(position = "dodge") + labs(title="Total Number of Ride by Month", x = "Month", y = "Number of Rides") + theme(axis.text.x = element_text(angle = 60, hjust = 1))
```


### Save file for analysis in Tableau
```{r}
write.csv(all_trips_v2, file = "~/Downloads/all_trips_v2.csv", row.names = FALSE)
```


### Check to see there are only 2 values for member_casual
```{r}
unique(na.omit(all_trips_v2$member_casual))
```
### Table results to see count of member_casual and create a data frame
```{r}
table(all_trips_v2$member_casual)
```


```{r}
df <- data.frame(
  type = c("Member", "Casual"),
  count = c(1678177, 1256371)
)
```

### Convert to percentages
```{r}
library(scales)
df$percent <- df$count / sum(df$count)
```

```{r}
df_new <- df %>%
  mutate(percent = count / sum(count)) %>%
  select(type, percent)
```


### Create a pie chart to depict the relationship between member_casual
```{r}
library(dplyr)
library(scales)

df <- data.frame(
  type = c("Member", "Casual"),
  count = c(1678177, 1256371)
)

df$percent <- df$count / sum(df$count)

df_new <- df %>%
  select(type, percent)
  
  

ggplot(df, aes(x="", y=percent, fill=type)) +
  geom_bar(stat="identity", width=1, color="white") +
  coord_polar("y", start=0) +
  theme_void() +
  theme(legend.position = "bottom") +
  scale_fill_manual(values=c("#F8766D", "#00BFC4"), 
                    labels=paste(df_new$type, scales::percent(df_new$percent)),
                    breaks=df_new$type)+ 
  labs(title="Member vs Casual Rider Percentages")
```


### See the total number of rides from dataset 
```{r}
library(ggplot2)
total_rides <- all_trips_v2 %>% 
  group_by(member_casual) %>% 
  summarize(count = n())

print(total_rides)
```

### Create a bar chart to show relationship between Total Number of Rides by Rider Type
```{r}
ggplot(total_rides, aes(x = member_casual, y = count, fill = member_casual)) +
  geom_col() +
  theme_bw() +
  labs(x = "Rider Type", y = "Total Number of Rides", fill = "Rider Type") +
  ggtitle("Total Number of Rides by Rider Type")
```


**Act**

**How do annual members and casual riders use Cyclistic Bike Share differently?**

 * Based on my analysis in R, casual riders have the highest ride duration every day of the week. Although casual riders hold the highest duration of ride each day, they only hold the highest total rides by day on weekends. **This data suggests that casual riders predominantly use the bike share system on the days of Saturday and Sunday**. 

 * Looking at the plot of "Total Number of Ride by Month", we can see that between the months of **July** and **August** there is the highest total number of rides for members and casual riders. 
 
 * After the month of August, the total number of casual rides by month drops dramatically. This rapid change in ridership activity could be due to warm temperatures dropping.
 

**Marketing Objectives**
 
 * To maximize the amount of casual riders retained through the end of the year, **create digital marketing plans to be implemented during the months of July to August**. These three months have the highest total number of rides by month consistently. Retaining riders through the year could help make Cyclistic Bike Share a riders primary source of transportation in downtown areas. 
 
 * Create **specials or promotions** to target casual riders during these peak months and help retain members during the slow season. 
 
 * Increase the number of available bikes in popular spots during these peak months. 
 

**Final Thoughts** 

Further analysis will be needed to see if there is a relationship between weather and decreased number of rides per month after August. There is also interest in analyzing member use to see what members use their rides for. This could be commuting to work, groceries,or leisure. This analysis will make it possible to draw conclusions to the drop in rides per month between the months of September to February. Further analysis will help to make further recommendations. 





